% This must be in the first 5 lines to tell arXiv to use pdfLaTeX, which is strongly recommended.
\pdfoutput=1
\documentclass[11pt]{article}

% Change "review" to "final" to generate the final (camera-ready) version.
\usepackage[review]{acl}

\usepackage{times}
\usepackage{latexsym}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{inconsolata}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}

\title{Can Large Language Models Count in Legal Czech? \\
Subject--Predicate Distance as a Probe of LLM Reliability for \\
Rule-Based Legislative Drafting Support}

\author{Anonymous submission}

\begin{document}
\maketitle

\begin{abstract}
Tools that support legislative drafting increasingly need to check \emph{structural} rules---limits
on paragraph counts, on cross-reference nesting, or on other measurable properties of a sentence.
We ask whether large language models (LLMs) can serve as the engine for such rules, using a narrow,
objectively scoreable proxy: identifying the main subject (\emph{podm\v{e}t}) and predicate
(\emph{p\v{r}\'{i}sudek}) of a Czech legal sentence and counting the linear word-distance between
them. We evaluate two family-matched open model pairs (Llama-3.1-8B / Llama-3.3-70B and
Qwen-2.5-7B / Qwen-2.5-72B) under three prompting regimes---raw text, self-generated CoNLL-U, and
gold CoNLL-U supplied---on a stratified sample of 640 sentences from the KUK~1.0 corpus, each run
three times at temperature~0 (23{,}040 generations). Structure helps and larger models are better,
but no configuration is reliable: the best condition---the 72B model \emph{handed} the gold
parse---is still only 65\% exact. A decomposition shows the residual failure is \emph{counting},
not structure-reading: given the parse the large models name the correct subject and predicate
83--87\% of the time, yet even when they name both words correctly they still miscount the distance
28--36\% of the time; the small models miscount almost always. The instability is
severe (up to 56\% of triplets vary run-to-run even at temperature~0 with a pinned provider) and
majority voting does not fix it. A parser-quality control against a human gold treebank shows the
silver oracle agrees with humans 99.3\% on the labels, so these failures are real. We conclude that
structural drafting rules should be implemented as deterministic extractors over a dependency parse
($\approx$99\% accurate and zero-variance) rather than as LLM prompts.
\end{abstract}

\section{Introduction}

Software that assists in drafting or reviewing legislation is beginning to encode explicit
\emph{rules} for text quality. Some of these rules are semantic (``is this provision unnecessarily
convoluted?''); others are \emph{structural} and involve exact measurement---e.g. a limit on the
number of paragraphs per section, on the nesting depth of cross-references, or on other
complexity metrics derived from sentence structure. A natural implementation question is whether a
single LLM prompt can serve as the engine for both kinds of rules, or whether structural rules
require a different mechanism.

Independently, a growing body of work shows that LLMs are unexpectedly unreliable at \emph{exact
counting}. Performance swings by large margins with tokenization \citep{zhang2024counting}, models
fail at seemingly trivial character counts \citep{fu2024count}, and counting has been proposed as a
minimal probe of rule-following reliability \citep{dai2026rulefollowing}. If LLMs cannot reliably
count in a controlled setting, they are a poor foundation for counting-based drafting rules.

We make this concrete with a deliberately narrow task in Czech legal text: given a sentence,
identify its main-clause subject and predicate and compute the number of words that lie between
them. The task is a proxy---not a drafting rule itself---but it is objectively scoreable, it is
directly tied to an established syntactic-complexity measure (dependency distance;
\citealp{gao2024dependency}), and it isolates counting from heavy semantics. We use it to answer
three questions: (RQ1)~can LLMs identify the subject/predicate pair? (RQ2)~can they count the
distance? (RQ3)~does supplying explicit syntactic structure (a CoNLL-U parse) help, and does the
effect depend on model size?

Our contributions are: (i)~a fully reproducible, provider-pinned evaluation of four open models
across three prompting regimes on 640 stratified legal sentences; (ii)~a decomposition separating
\emph{role identification} from \emph{counting}, which reframes the ``can LLMs count'' question;
(iii)~a quantification of run-to-run instability and a demonstration that majority voting does not
remedy it; (iv)~a parser-quality control against a human gold treebank establishing that the
silver oracle is trustworthy on the labels; and (v)~a concrete recommendation for structural
drafting rules. Code, prompts, the sampled evaluation set, and all model traces are released.

\section{Related Work}

\paragraph{LLM counting and rule-following.} \citet{zhang2024counting} show tokenization alone
causes large accuracy swings on counting; \citet{fu2024count} document persistent failures on
character counting. \citet{dai2026rulefollowing} frame stable counting as a minimal probe of
rule-following reliability. Our task instantiates this probe on natural legal text with a
linguistically motivated target.

\paragraph{Structured input for parsing-like tasks.} \citet{ginn2025llm} show that supplying
symbolic/in-context rules improves LLM zero-shot dependency parsing, with the benefit shrinking as
more examples are added. Our regimes 2 and 3 (self-generated vs.\ supplied CoNLL-U) test the analogous
question for a structural measurement task.

\paragraph{Dependency distance as complexity.} Linear distance between a syntactic head and its
dependent is an established metric of syntactic complexity and working-memory load
\citep{gao2024dependency,liang2007dependency}. Subject--predicate distance is arguably the single
most load-bearing instance, which grounds our metric as more than an ad-hoc token count and ties it
to the comprehensibility concerns of legislative drafting.

\paragraph{Czech legal NLP resources.} We use the KUK~1.0 corpus of Czech legal/administrative text
\citep{hladka2024kuk}, parsed with UDPipe~2 \citep{straka2018udpipe} into Universal Dependencies
\citep{nivre2020ud}, and the Czech Legal Text Treebank \citep{kriz2018cltt} (via its UD conversion)
as a human-gold parser-quality reference.

\section{Task and Data}

\subsection{Task formalization}
Given a sentence, let $p$ be the position of the main-clause subject and $q$ the position of the
predicate. The distance is
\begin{equation}
d = |p - q| - 1,
\end{equation}
the number of tokens strictly between them in linear order (not tree depth). We operationalize the
\textbf{subject} as the single \texttt{nsubj}/\texttt{nsubj:pass} dependent of the sentence root,
and the \textbf{predicate} as the head of that subject---i.e.\ the root---whatever its part of
speech. This handles copular clauses correctly and matches the head--dependent pair that the
dependency-distance literature measures: in \emph{``Z\'{a}kon je \'{u}\v{c}inn\'{y}''} (``The law is
effective'') the predicate is the nominal/adjectival part \emph{\'{u}\v{c}inn\'{y}}, with the copula
\emph{je} recorded separately.\footnote{This choice is provisional pending confirmation by a
linguist; the copula index is logged so the alternative is recoverable without re-running.}

Crucially, distance is counted in \textbf{whitespace-delimited words}, not CoNLL-U tokens. This is
the human-meaningful unit and the only unit a model can count on raw text without being handed a
tokenization standard; punctuation glued to a word and Czech split-tokens (e.g.\ \emph{aby}
$\rightarrow$ \emph{aby}+\emph{by}) each count as one word. We map UDPipe's tokens to words using
the CoNLL-U character offsets, so gold and model answers are commensurable in every regime.

\subsection{Data}
KUK~1.0 \citep{hladka2024kuk} contains four sub-corpora of Czech legal/administrative text
(ombudsman opinions, Frank Bold materials, ombudsman information flyers, and mixed advice texts),
with silver UD parses from UDPipe~2 (\texttt{czech-pdt-ud-2.15}) and named-entity tags from NameTag~3
\citep{strakova2019nametag}. It comprises 6{,}275 CoNLL-U files and 910{,}373 sentences. We treat
UDPipe's parse as the silver gold; \S\ref{sec:noisefloor} quantifies how far this can be trusted.

\paragraph{Filtering.} We keep sentences with exactly one root, exactly one root-attached
\texttt{nsubj} (no coordinated/clausal subjects, no multiple finite clauses), and word length in
$[5,80]$. This retains \textbf{415{,}462 / 910{,}373 (45.6\%)} sentences, consistently across
sub-corpora---itself a descriptive statistic about how often ``simple'' subject--predicate
structures occur in legal Czech.

\paragraph{Evaluation sample.} From the filtered pool we draw a stratified sample of
\textbf{N=640}, crossing four sentence-length buckets with four distance buckets ($4\times4$) and
allocating an \emph{equal} 40 sentences per stratum. This deliberately over-samples the long/far
tail (population median distance is~1; sample median is 4.5), where difficulty concentrates. The
draw is balanced across the four sub-corpora (160 each) so results are not implicitly a claim about
one genre, and is fully seeded for reproducibility.

\section{Experimental Setup}

\paragraph{Regimes.} Each sentence is presented under three regimes: (1)~\textbf{raw text}---the
model reads the sentence and returns subject, predicate, and distance; (2)~\textbf{self-CoNLL-U}---the
model first emits its own CoNLL-U parse, then the answer; (3)~\textbf{CoNLL-U given}---the model is
handed UDPipe's parse and only needs to locate \texttt{nsubj}, follow its head, and count. All
prompts are in Czech with two few-shot examples (one verbal, one copular) identical across regimes.

\paragraph{Models.} We use two family-matched pairs so size is manipulated within an architecture:
Llama-3.1-8B and Llama-3.3-70B \citep{dubey2024llama}, and Qwen-2.5-7B and Qwen-2.5-72B
\citep{qwen2025}. All are served at fp8 through a single pinned provider per model, so quantization
is held constant and the run is reproducible.

\paragraph{Protocol and metrics.} Each (sentence $\times$ regime $\times$ model) cell is run three
times at temperature~0, one sentence per call---23{,}040 generations. We report: format validity
(a parseable answer); \emph{role identification} (subject/predicate word-index match); \emph{distance}
exact-match, $\pm1$, $\pm2$, and MAE; and \emph{consistency} (fraction of 3-run triplets that are
identical). Full request/response traces, including the provider-reported per-call cost, are stored;
the total run cost was USD~7.73.

\section{Results}

\begin{table}[t]
\centering\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{llrrr}
\toprule
\textbf{Model} & \textbf{Regime} & \textbf{Exact} & \textbf{$\pm1$} & \textbf{Pos.} \\
\midrule
Llama-8B  & raw          & 8.0  & 29.5 & 5.2  \\
          & self-CoNLL-U & 8.2  & 40.7 & 18.0 \\
          & CoNLL-U given& 4.2  & 51.6 & 35.4 \\
\addlinespace
Qwen-7B   & raw          & 7.4  & 20.3 & 2.8  \\
          & self-CoNLL-U & 12.2 & 42.8 & 21.3 \\
          & CoNLL-U given& 23.9 & 51.5 & 32.1 \\
\addlinespace
Llama-70B & raw          & 20.9 & 42.7 & 10.6 \\
          & self-CoNLL-U & 26.1 & 43.8 & 20.1 \\
          & CoNLL-U given& \textbf{59.6} & 67.5 & 50.8 \\
\addlinespace
Qwen-72B  & raw          & 25.3 & 49.4 & 15.5 \\
          & self-CoNLL-U & 39.1 & 56.3 & 32.0 \\
          & CoNLL-U given& \textbf{64.9} & 71.1 & 51.6 \\
\bottomrule
\end{tabular}
\caption{Distance exact-match, $\pm1$-tolerance, and \textbf{Pos.}\ (both subject and predicate
reported at their correct word position) (\%), by model and regime, over 640 sentences $\times$ 3
runs. \S\ref{sec:decomp} separates \emph{naming} the right words from \emph{positioning} them.
Format validity was $\geq$93\% throughout ($\geq$99\% except Llama-8B self-CoNLL-U).}
\label{tab:main}
\end{table}

Table~\ref{tab:main} presents the main results. We organize the discussion around four hypotheses.

\paragraph{H1: structure helps.} Exact-match rises from raw to CoNLL-U-given for three of four
models, and strongly for the large ones: Llama-70B $20.9\!\rightarrow\!59.6$ and Qwen-72B
$25.3\!\rightarrow\!64.9$. Using paired McNemar tests on the large models, every pairwise regime
difference is significant, and raw~vs.~CoNLL-U-given is overwhelming ($p<10^{-45}$). The value lies
almost entirely in \emph{being handed} the parse, not in generating it (self-CoNLL-U helps only
modestly, $p\approx10^{-8}$ to $0.009$).

\paragraph{H2: larger is better.} Within both families and every regime, the larger model wins by a
wide margin (e.g.\ CoNLL-U-given: 59.6 vs.\ 4.2 for Llama; 64.9 vs.\ 23.9 for Qwen).

\paragraph{H3: does structure \emph{compensate} for size?} No---if anything it \emph{widens} the
gap. The raw-text size gap is $\approx$13--18 points; with CoNLL-U given it is $\approx$41--55
points. Structure lets large models convert the parse into a correct count; the small models cannot.

\paragraph{H4: degradation with distance.} Exact-match falls monotonically with the gold distance.
Pooled over models, the CoNLL-U-given rate across distance buckets $\{0\text{-}1,2\text{-}4,5\text{-}9,10{+}\}$
is $48\!\rightarrow\!51\!\rightarrow\!39\!\rightarrow\!15$; raw text is $26\!\rightarrow\!22\!\rightarrow\!11\!\rightarrow\!3$.
The long tail---exactly the adversarial case for legal Czech---collapses in every regime.

\subsection{Finding the words vs.\ counting the span}
\label{sec:decomp}

The best condition---Qwen-72B handed the gold parse---is still only 64.9\% exact. It is tempting to
read this as the model failing to \emph{read} the supplied structure, but the residual failure is
mostly \emph{counting}. We separate two abilities: \emph{naming} the correct subject and predicate
words, and \emph{counting} the distance between them (Table~\ref{tab:pure}).

\begin{table}[t]
\centering\small
\setlength{\tabcolsep}{5pt}
\begin{tabular}{lrrr}
\toprule
\textbf{Model} & \textbf{Names} & \textbf{Dist.} & \textbf{Exact $\mid$} \\
\textbf{(CoNLL-U given)} & \textbf{both} & \textbf{exact} & \textbf{both named} \\
\midrule
Qwen-72B  & 82.7 & 64.7 & 72.3 \\
Llama-70B & 86.7 & 59.6 & 63.8 \\
Qwen-7B   & 41.0 & 23.5 & 38.4 \\
Llama-8B  & 45.5 &  4.2 &  0.1 \\
\bottomrule
\end{tabular}
\caption{With the parse supplied: how often the model \emph{names} both the correct subject and
predicate words, its overall distance exact-match, and the exact-match \emph{restricted to} the
cases where it named both words correctly (all \%). Even given the parse \emph{and} the right two
words, the large models still miscount $\approx$28--36\% of the time; Llama-8B essentially never
counts correctly.}
\label{tab:pure}
\end{table}

Given the parse, the large models \emph{name} the right subject and predicate most of the time (both
correct 83--87\%), yet their distance is exact only 60--65\%. Even restricting to generations where
they named \emph{both} words correctly, the distance is exact only 63.8\% (Llama-70B) and 72.3\%
(Qwen-72B): handed the parse \emph{and} the right two words, the large models still miscount the span
$\approx$28--36\% of the time. Llama-8B essentially never produces the correct distance even when it
named both words (0.1\%).

The error is specifically in \emph{counting word positions}. The large models report the correct
predicate word 93--97\% of the time but its correct linear position only ${\approx}55\%$ of the time
(subject: named 87--88\%, positioned 73--75\%); the endpoint deeper in the sentence---usually the
predicate---is mis-positioned far more often, the signature of a count that drifts with length. In a
typical failure the model names the right words (e.g.\ \emph{provozovatel}, \emph{hrad\'{i}}) but
places them at positions 30 and 17 instead of 25 and 14, giving distance 12 instead of 10. In short,
with the structure supplied, identifying \emph{which} words are subject and predicate is largely
solved; what remains unreliable---even for the 72B model---is the counting itself.

\subsection{Instability}
\label{sec:instab}

Even at temperature~0 with a pinned provider and quantization, the three runs disagree on
\textbf{36\%} (raw text) and \textbf{56\%} (self-CoNLL-U) of sentence$\times$model triplets;
CoNLL-U-given is far more stable (11\%). Two points matter for deployment. First, the instability
\emph{tracks difficulty}: for raw text the fraction of varying triplets grows with distance
($29\%\!\rightarrow\!41\%$). Second, and decisively, \textbf{majority-vote-of-three does not improve
accuracy}---it equals the single-run rate in every cell (e.g.\ Qwen-72B raw $25.3\!\rightarrow\!24.8$;
Llama-70B CoNLL-U-given $59.6\!\rightarrow\!60.0$). When the model varies, it flip-flops between
\emph{wrong} answers, so self-consistency or resampling will not deliver repeatable rule outcomes.
A \emph{consistent} answer can also be consistently \emph{wrong}: Llama-8B given CoNLL-U is 80\%
consistent but 4\% correct---reproducibility is not correctness.

\subsection{Is the silver oracle trustworthy?}
\label{sec:noisefloor}

Because we score against UDPipe, we re-parse 740 human-gold sentences from the Czech Legal Text
Treebank \citep{kriz2018cltt} with the \emph{same} UDPipe model and compare. Where UDPipe commits to
a comparable single-clause parse (295/740), it agrees with the human gold on subject (100\%),
predicate (99.3\%), and \emph{distance-exact} (99.3\%), MAE~0.14. The silver \emph{labels} are thus
highly reliable on legal Czech, and the model failures above are real, not oracle noise. The silver
noise that does exist is in \emph{sentence segmentation}: for the remaining 445 sentences UDPipe
split the gold sentence into two or more, disagreeing about where a sentence ends rather than about
its subject and predicate.

\section{Discussion: implications for drafting tools}

Two results together are decisive for how structural rules should be built. First, the LLM's weak
link is the counting itself: even handed the gold parse \emph{and} having named the correct subject
and predicate, the best model still miscounts the distance $\approx$28\% of the time
(\S\ref{sec:decomp})---whereas computing the distance from two identified endpoints is deterministic
arithmetic, exact in code. Second, our noise-floor shows a dependency parser identifies those
endpoints $\approx$99\% correctly on legal Czech (\S\ref{sec:noisefloor}). Hence a \textbf{deterministic
extractor over the UD parse would be $\approx$99\% accurate and zero-variance}, against $\leq$65\%
and highly unstable for the best LLM prompt. Standard LLM mitigations---temperature~0, provider
pinning, majority voting---demonstrably do \emph{not} yield repeatable counts (\S\ref{sec:instab}).

We therefore recommend that legislative-drafting tools \textbf{triage each rule}: structural /
countable rules (paragraph counts, reference nesting, dependency-distance-style complexity) should be
implemented as deterministic functions of a parse, \emph{not} as LLM prompts; ``LLM~+~structured
context'' is warranted only for rules requiring semantic judgment a parser cannot express. The one
caveat that carries over to any method is sentence segmentation, which on legal Czech is
parser-dependent and must be treated as a fallible, first-class step upstream of any counting rule.

\section{Conclusion}

On a narrow, objectively scoreable proxy---subject--predicate word-distance in Czech legal text---no
LLM configuration we tested is reliable, the best (a 72B model handed the gold parse) reaching only
65\% exact. Larger models read the structure reasonably well but still miscount the span
$\approx$30\% of the time even when handed the parse and having named the correct endpoints; smaller
models cannot count at all; and the run-to-run instability cannot be tuned away. Because a
dependency parser identifies the endpoints $\approx$99\% correctly and the count is then
deterministic, structural drafting rules are best served by deterministic extraction over a parse
rather than by prompting an LLM.

\section*{Limitations}

The evaluation is silver-standard: we measure agreement with UDPipe, not linguistic truth. We
mitigate this with a human-gold parser-quality control (\S\ref{sec:noisefloor}) showing 99.3\%
label agreement, but the segmentation disagreement it surfaces is a genuine limitation for
end-to-end deployment. The operationalization of ``predicate'' as head-of-subject is a defensible
but provisional choice awaiting linguist confirmation. We test one prompt formulation per regime
and do not perform prompt-engineering sweeps; results may shift with different phrasings. The four
models are all open-weight and $\leq$72B; whether the counting bottleneck shrinks at frontier
scale is an open question we did not test. Finally, success or failure on
this proxy predicts, but does not prove, (un)reliability on other counting-type rules (paragraph
counts, reference depth); we frame it as a probe, not a general claim about ``LLMs and counting.''

\bibliography{custom}

\end{document}
